Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-10 paper.tex
Editing imputation section and figure

\subsection{Imputation of Pango lineage for non-sample nodes}
% FIXME the caption and the text are basically saying the same thing here. Can
% we streamline, and maybe give a bit more detail in the text about iteration,
% and the concrete implementation?
\begin{figure} \centering
\includegraphics[width=.7\textwidth]{figures/imputation.pdf}
\caption{\label{fig:imputation}A schematic of the two-step procedure to impute
Pango lineage for inserted, non-sample nodes. Here, three nodes (question
marks) have unknown Pango lineage (A). Because the parent of node 1 is
available (i.e., L2) and there is no mutation occurring on the edge connecting
the two nodes, the Pango lineage of the parent is directly propagated to node 1
(B). However, if there is a mutation (red X’s), then the Pango lineage is
inferred by comparison to the set of lineage-defining mutations from
COVID-CG\citep{Chen2021-zc}, as in case (C). Finally, the inferred status is
propagated to any non-mutant child nodes, as in case (D).
}
\end{figure}
For many applications it is helpful to infer the Pango lineage status
of the internal ARG nodes estimated by \texttt{sc2ts}.
% Initially, only sample nodes have Pango lineage assignments.
% For analysis of recombinant parents, it is also helpful to assign
% Pango lineages to non-sample nodes.
We do this using a two-step procedure (Figure~\ref{fig:imputation}).
Initially, only sample nodes have Pango lineage assignments.
For a given non-sample node $u$, if the Pango lineage of its parent is
available and there are no intervening mutations,
then we impute that $u$'s Pango lineage is the same as its parent.
Otherwise, the Pango lineage of the
non-sample node is inferred by matching its full set of mutations against the
Pango lineage-defining mutations (based on 90\% consensus of the sequences
analysed) from the COVID-CG website \citep{Chen2021-zc};
\url{https://covidcg.org/}; accessed on November 04, 2022).

[TODO how accurate is this approach? We should probably discuss it somewhere
in the Results, and refer back from here? ]

>>>>

\subsection{Imputation of Pango lineage for non-sample nodes}
% FIXME the caption and the text are basically saying the same thing here. Can
% we streamline, and maybe give a bit more detail in the text about iteration,
% and the concrete implementation?
\begin{figure} \centering
\includegraphics[width=.7\textwidth]{figures/imputation.pdf}
\caption{\label{fig:imputation}A schematic of the iterative procedure to impute
Pango lineage for inserted, non-sample nodes. Here, three nodes (question
marks) have unknown Pango lineage (A). The lineage for node N1 can be directly 
copied from its parent (L2). The lineage for node N2 must be inferred from 
that of the parent (L2) plus the lineage-defining mutation (red X) 
on the connecting edge. The lineage for node N3 can then be copied from 
that of node N2.}
\end{figure}
While the Pango lineages of sample nodes are imported directly from 
GISAID metadata, the lineage status of internal nodes inserted into 
the \texttt{sc2ts} ARG must be imputed. We do this using 
the list of lineage-defining mutations (based on 90\% consensus 
of the sequences analysed) from the COVID-CG website \citep[][\url{https://covidcg.org/}; accessed on 4 November 2022]{Chen2021-zc}.

For a given non-sample node $u$, if the Pango lineage of its parent or 
one of its children is already known, and there are no lineage-defining mutations
on the connecting edge, then $u$ copies this Pango lineage exactly.
Otherwise, the lineage for $u$ is inferred by 
matching its full set of mutations against the COVID-CG list. 
We apply these two steps to the internal nodes of the ARG iteratively,
as illustrated in Figure~\ref{fig:imputation}, until all internal nodes 
are assigned a lineage (where possible---note that sometimes a lineage cannot
be assigned to the children of a recombination node).

This method is fast, as the lineages for most of the internal nodes can be 
imputed by copying from the surrounding nodes (66\% of nodes for the Long ARG, 
80\% of nodes for the Wide ARG). The accuracy depends on
the quality of the list of lineage-defining mutations; we find that when
masking the lineages of sample nodes one-by-one and re-imputing these using
the remaining information results in 98\% (resp.~99\%) of lineages being
imputed correctly.
